
---
output: github_document
---

# MBIO 612 Practical 3: Data Visualization
## Kathryn Van Artsdalen

```{r}
library(tidyverse)
```

**Question 1**

* Provide at least two methods to view the number of rows and the number of columns in the `mpg` dataset

```{r}
data(mpg)
# First way is to use nrow and ncol functions to count the number of rows and columns, respectively
ncol(mpg) 
nrow(mpg)

#Second way by calling dimensions of the mpg dataset
dim(mpg)

# A third way is to make a vector of the column names and use the length function to count the number of characters/objects in the vector. Repeat for row names.
mpg_columns <- colnames(mpg)
length(mpg_columns)
mpg_rows <- rownames(mpg)
length(mpg_rows)
```

**Question 2**

* Run the following code:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

* What is wrong with the code. 
* Fix it so that the plot is displayed correctly.

The code uses the color aesthetic inside the aes() but does not group based on a category, such as class, so there is no differentiation between variables. There is no way to scale "blue" since "blue" is not a variable name. Color = "blue" actually returns data points that are red, and does not convey any additional information about them. To change the color of all points to blue, color must be listed outside aes() so it treats all variables the same.

```{r}
# To color by a variable (such as class)
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

```{r}
# To color so all points are blue
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

**Question 3**

* Change the code below so that mpg$class_numeric will be of type integer (int) rather than double (dbl)
```{r}
# mpg$class_numeric = sapply(mpg$class, convert_class_int, USE.NAMES = FALSE)
```
* Inspect the result to make sure your solution worked

```{r}
mpg$class_numeric <- as.integer(factor(
  mpg$class,
  levels = c("2seater", "subcompact", "compact", "midsize", "minivan", "suv", "pickup"),
  ordered = TRUE
))
```

**Question 4** 

* Use a scatter plot to plot the variable `cty` (`x`-axis) and `hwy` (`y`-axis) of the `mpg` dataset
* Add another geom that shows the liner regression between `x` and `y`
  * The method for computing linear regression should be a linear model `lm` instead of the default (`loess`)
  * Set the line color to `red`
  * color the points by `class` 
  * set the points size to 6

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy))+
  geom_point(mapping = aes(color = class), size = 6)+
  geom_smooth(method = "lm", color = "red")+
  labs(title = "City vs Highway Fuel Consumption by Car Class", x = "City Mileage in mpg", y = "Highway Mileage in mpg")+
  theme_classic()
```

**Question 5** 

* You noticed in the previous plot that there exist in the dataset subcompacts that has a city fuel consumption of 15 miles per gallon.
* Extract the manufacturer and model of these subcompacts
  * results should only have these columns

```{r}
mpg[which(mpg$cty == 15 & mpg$class == "subcompact"), c("manufacturer", "model")]
```

**Question 6**

* The following generates a warning
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

* How can you fix that the code so all classes are displayed
* Hint: you can manually specify your own values for the shapes to use using `scale_shape_manual` geom.  You can select any of the [following shapes](https://ggplot2.tidyverse.org/articles/ggplot2-specs_files/figure-html/unnamed-chunk-7-1.png)

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))+
  scale_shape_manual(values = c(0,1,5,8,15,16,17))+
  labs(title = "Engine Displacement and Highway Mileage by Car Class",
       x = "Engine Displacement in litres",
       y = "Highway Mileage in mpg")+
  theme_classic()
```

**Question 7**

* Given the following two vectors

```{r}
x <- c(3.385,0.48,1.35,465,36.33,27.66,14.83,1.04,4.19,0.425,0.101,0.92,1,0.005,0.06,3.5,2,1.7,2547,0.023,187.1,521,0.785,10,3.3,0.2,1.41,529,207,85,0.75,62,6654,3.5,6.8,35,4.05,0.12,0.023,0.01,1.4,250,2.5,55.5,100,52.16,10.55,0.55,60,3.6,4.288,0.28,0.075,0.122,0.048,192,3,160,0.9,1.62,0.104,4.235)
y <- c(44.5,15.5,8.1,423,119.5,115,98.2,5.5,58,6.4,4,5.7,6.6,0.14,1,10.8,12.3,6.3,4603,0.3,419,655,3.5,115,25.6,5,17.5,680,406,325,12.3,1320,5712,3.9,179,56,17,1,0.4,0.25,12.5,490,12.1,175,157,440,179.5,2.4,81,21,39.2,1.9,1.2,3,0.33,180,25,169,2.6,11.4,2.5,50.4)
```
* Combine the two vectors into a data frame called `sizeMass`
  * Name the columns containing the vectors `x` and `y` as `size` and `mass` respectively 

* Use the code below to plot the data
  * If you followed the above instructions the code should work without modification

```{r}
sizeMass <- data.frame(size = x, mass = y)
```

```{r}
ggplot(sizeMass, aes(size, mass)) +
  geom_point() +
  geom_smooth(method = "lm", color = "red", se = FALSE)

```  


**Question 8** 

* In the above, the data is "squashed" because the data range varies substantially across observations. One way to fix this is by computing the log of the data. We will do that using two different approaches

1. add two new columns called them `size_log10` and `mass_log` that represent the log 10 values of the `size` and `mass` columns. Then, reuse the plotting code above by replacing `size` and `mass` with `size_log10` and `mass_log10` respectively
  * Hint `log10()` can take as input a single value or a vector

2. Use the `scale_x_log10()` and `scale_y_log10()` to set different scales for the `x` and `y` aesthetics
  * Simply add these transforms to the initial snippet of code prior to plotting `geom_point` or `geom_smooth`

```{r}
# Method 1
sizeMass.log10 <- cbind(sizeMass, size_log10 = log10(x), mass_log10 = log10(y))
ggplot(sizeMass.log10, aes(size_log10, mass_log10)) +
  geom_point() +
  geom_smooth(method = "lm", color = "red", se = FALSE)

```

```{r}
# Method 2
ggplot(sizeMass, aes(size, mass)) +
  scale_x_log10()+
  scale_y_log10()+
  geom_point() +
  geom_smooth(method = "lm", color = "red", se = FALSE)

```